Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-13 text/manuscript-full-KS.qmd
front end revisions

<!--Reactivity, discipline, and authority are three interrelated themes that describe the effects of quantification. Reactivity is the idea that salient quantitative measures cause people and organizations to change their behavior. Quantification disciplines actors to react in particular ways. Quantification changes decision-making power authority, often weakening the discretion of local decision-makers. For instance, scholarship about U.S. News & World Report (USNWR) law school rankings demonstrate the effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Law school rankings were developed as a means of informing prospective students about the relative quality of different programs. Rankings affect (reactivity) where students apply and enroll [@RN5003]. Rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- used rankings to make decisions that had large consequences for schools \ldots [then] schools felt pressured to take them seriously" [@RN5007, p. 415]. Rankings disciplined the behavior of law schools. For example, characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid [@RN5003]. Rankings also weakened the authority of admissions personnel to make independent decisions about class size and which applicants to admit [@RN5003]. <!--
In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.    -->

<!--The analysis of UK school "league tables" by @RN5005 shows how quantification can be a mechanism of social reproduction. In 1992, the UK government began publishing school league tables, which ranked schools based on student performance on national exams taken at age 16. League tables facilitate making evaluative comparisons between schools regardless of geographic proximity. They also discipline households and schools to conceive of school quality in terms of test scores, which are substantially a function of the class composition of schools. Using Census data measured at the local authority level (similar to a U.S. county), @RN5005 find that localities with higher performing schools experienced growth in the share of managerial/professional households following the adoption of league tables and a decline in the share of working-class households. Using longitudinal survey data, managerial/professional households were more likely to move to localities with higher ranked schools following the introduction of school league tables. The introduction of league tables contributed to social reproduction because professional/managerial households were more aware of these consumer-facing metrics and had resources to respond by moving to more expensive neighbhorhoods, near higher performing schools. THIS IS RATCHET EFFECT STUFF. I CUT IT BECAUSE YOU HAVEN'T INTRODUCED THE IDEA OF RATCHET EFFECT
Schools that performed well on initial league tables tended to be schools with a high share of professional/managerial families. Therefore, quantification takes a snapshot of existing social stratification and that snapshot becomes a power unto itself, amplifying the effect of previous strctural inequality on subsequent structural inequality.
-->

```{=html}
<!-- THIS IS THE CUT TRANSITION PARAGRAPH
In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].
-->
```

@@

<!-- SAT test-takers were categorized into one of four different market segments -- local, in-state, regional, and national -- based on SAT score-sending behavior. For a given student, each college that receives a score from the student can be defined as "local" (college located in the the same Geomarket as the student), "in-state" (same state but different Geomarket), "regional" (same region but different state), or "national" (different region). In turn, a test-taker is categorized in the "local" market segment if they submit more SAT scores to local institutions than they do to in-state, regional, or national institutions. An "in-state" student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions, etc. 
A "regional" student submits more SAT scores to regional institutions than they do to local, in-state, or national institutions. A "national" student submits more SAT scores to national institutions than they do to local, in-state, or regional institutions.
-->

The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family in come greater than \$35,000, percent with both parents having a BA, etc.[^5] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.[^6] Taken together, the Market Segment Profile shows a college which Geomarkets possess attractive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

@@

@RN5006 argue that quantification has effects to the extent that stakeholders care about the numbers. Market research suggests that EPS was highly salient. @RN5021 reports that in 1995, 37% of 4-year publics and 49% of 4-year privates used EPS, while 41% of 4-year publics and 16% of 4-year privates used ACT's market analysis service product.

EPS software may "discipline" [@RN5007] colleges to approach recruiting in a manner consistent with the Market Segment Model. Drawing from promotional literature [e.g., @RN5019; @RN1621], @RN4980, and background conversations with enrollment managers, the practical purpose of EPS was to inform the "travel schedule" of admissions recruiters. EPS promotional material and user guides encourages users to begin by identifying which Geomarkets they will recruit from.[^9] Second, EPS users decide which high schools they will visit within selected Geomarkets. Based on the principle of homophily, the Market Segment Model suggests that selective colleges should target Geomarkets with large numbers of affluent, college educated households, while low-income communities are left to local four-year and community colleges.

@@

EPS may also weaken the authority of local decision-makers. @RN4982 sought to develop a concrete, data-driven framework -- The Market Segment Model -- that replicates the aggregate knowledge of local admissions officers. Once this knowledge was quantified and commodified onto a CD-ROM, the local expertise of admissions officers becomes less valuable. The EPS product increases the ability of a college admissions leader -- working with College Board staff or an EM consultant -- to plan recruiting efforts centrally. In background conversations, EM professionals told us that EPS software enabled colleges to plan travel without relying on admissions officers having strong local knowledge of their territories. However, EPS software often recommended visiting the same sets of affluent, high-achieving high schools that were receiving visits from other colleges. We were told that savvier, quantitatively-adept admissions offices used EPS to visit schools that EPS recommended ignoring because there would be less competition for the good students who attended these high schools.

@@

Although we cannot gain access to EPS software, Geomarkets are fundamental to EPS software and (arguably) to the organization of recruiting territories for many colleges. Therefore, we are interested in the extent to which this building block is associated with race and class. Research question 1 asks, what is the socioeconomic and racial variation between Geomarkets and how does this variation change over time? We focus on metropolitan areas that are associated with three or more Geomarkets.

We expect substantial socioeconomic inequality between geomarkets in large metropolitan areas. Given the extent of class- and race-based residential segregation in the U.S., it would be surprising to not observe such inequality. Moreover, @RN4982 viewed demand for higher education as a function of class and developed Geomarkets with an eye towards identifying geographic areas that differed from one another in terms of class composition.

```{=html}
<!-- 
Additionally, @RN4982 created geomarkets in service of the Market Segment Model. The Market Segment Model argues that socioeconomic status drives whether students belong to the local, in-state, regional, or national market segment. In turn, geomarkets are described by the relative numbers of local, in-state, regional, or national students, which is substantially a function of socioeconomic status. The language of @RN4982 [pp. 11-12, quoted above] suggests that geomarket borders within large metropolitan areas may have been drawn in ways that separate affluent from less affluent communities. We present the following hypothesis:
-->
```
We also expect substantial racial inequality between geomarkets in large metropolitan areas. Interestingly, the Market Segment Model is explicitly based on socioeconomic stratification, but @RN4982 do not mention race. U.S. cities are characterized be extreme historic and contemporary residential racial segregation [@RN4779]. Structures built upon racialized structures are racialized structures [@RN4786]. Unless designers intentionally consider racial segregation, selection devices that categorize people based on geographic location are likely to reproduce historical race-based inequality in opportunity [@RN4975]. Second, although geomarket borders may have been drawn along class divides, a strong correlation exists between race and wealth [@RN4884]. Third, Geomarket borders may have been drawn in a way that follows the countours of racial segregation in residential housing. Examples include: the "South and South Central Los Angeles" geomarket (CA21); the "City of Oakland" geomarket (CA07), which is surrounded by the "Alameda County excluding Oakland" geomarket (CA08); and the "Wayne County Detroit" geomarket (MI01), which is surrounded by the "Detroit's Northern Suburbs" (MI02) and "Ann Arbor" (MI03) geomarkets.

**Geomarket Filter in Student Search Service**. <!-- By 1984, College Board Geomarkets were incorporated as a search filter in the College Board Student Search Service \[CITE\]. --> A student list contains the contact information of prospective students who meet the search filter criteria (e.g., test score, GPA) specified by the university. Student lists are the fundamental input for undergraduate recruiting campaigns because purchased names -- alongside prospects who reach out on their own -- constitute the set of prospects who receive subsequent recruiting interventions (e.g., mail, email) designed to push them toward the application and enrollment stages of the “enrollment funnel.” @RN4895 reports that 86% of public colleges and 87% of private colleges purchase student lists.[^10] Historically, dominant list vendors are college board and ACT, which derived student list data from test-takers, but the test-optional movement, advances in technology, and surging private equity investments have contributed to new sources of student list data [@list_biz].

@@

Because of the extent of residential segregation in the U.S., geographic borders are a commonly studied racialized input in scholarship about algorithmic bias [@RN4775; @RN4849; @RN4887]. Student list products offer many geographic filters. Some geographic filters are based on known geographic borders (e.g., zip code, county, CBSA, state). College Board also created geographic borders that subsequently became filters in the *Student Search Service*, for example the Geodemographic Segment filters.[^13] This manuscript focuses on Geomarkets. In their sample, @RN5012 found that 11% of student lists purchased by public research universities filtered on Geomarkets. Drawing from @RN4786, we argue that geomarkets satisfy the two criteria of racialized inputs. First, they are ostensibly race-neutral inputs in that neither @RN4982 -- nor subsequent EPS promotional material [e.g., @RN1621] -- mention race. Second, Geomarkets may be systematically correlated with race. For example, Geomarket "CA07" is "City of Oakland" and "CA08" is "Alameda Country, excluding Oakland." A university might purchase prospect contact information by filtering on CA08 in combination with additional filters, such as year of high school graduation and SAT test score. This article extends @RN5012 by examining how the racial composition of actual student list purchases changes if universities filter on particular Geomarkets.

@@

In both mailings and off-campus recruiting visits, @RN4982 recommend focusing on Geomarkets with large populations of students who are deciding between "institutions like your own" (p. 44). Selective colleges primarily draw from students in the "regional" and "national" segments. @RN4982 suggest that the student list purchases of selective colleges should focus on Geomarkets with large numbers of affluent, highly educated households. By contrast, the student list purchases of local state colleges and community colleges, which rely on enrollment from students in the "local" and "in-state," should focus on nearby Geomarkets with large numbers of low-SES students. Our analyses attempt to recreate the strategy recommended by @RN4982 by showing who is included/excluded when student list purchases filter on particular Geomarkets. Considering the correlation between race and wealth in segregated America, we suggest that a class-based list-buying strategy will increase racial stratification in college access. Furthermore, we anticipate that the students of color from affluent Geomarkets will tend to have college-educated parents and be enrolled in affluent, predominantly white schools.

>>>>

```{=html}
<!--Reactivity, discipline, and authority are three interrelated themes that describe the effects of quantification. Reactivity is the idea that salient quantitative measures cause people and organizations to change their behavior. Quantification disciplines actors to react in particular ways. Quantification changes decision-making power authority, often weakening the discretion of local decision-makers. For instance, scholarship about U.S. News & World Report (USNWR) law school rankings demonstrate the effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Law school rankings were developed as a means of informing prospective students about the relative quality of different programs. Rankings affect (reactivity) where students apply and enroll [@RN5003]. Rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- used rankings to make decisions that had large consequences for schools \ldots [then] schools felt pressured to take them seriously" [@RN5007, p. 415]. Rankings disciplined the behavior of law schools. For example, characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid [@RN5003]. Rankings also weakened the authority of admissions personnel to make independent decisions about class size and which applicants to admit [@RN5003]. <!--
In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.    -->
```{=html}
<!--The analysis of UK school "league tables" by @RN5005 shows how quantification can be a mechanism of social reproduction. In 1992, the UK government began publishing school league tables, which ranked schools based on student performance on national exams taken at age 16. League tables facilitate making evaluative comparisons between schools regardless of geographic proximity. They also discipline households and schools to conceive of school quality in terms of test scores, which are substantially a function of the class composition of schools. Using Census data measured at the local authority level (similar to a U.S. county), @RN5005 find that localities with higher performing schools experienced growth in the share of managerial/professional households following the adoption of league tables and a decline in the share of working-class households. Using longitudinal survey data, managerial/professional households were more likely to move to localities with higher ranked schools following the introduction of school league tables. The introduction of league tables contributed to social reproduction because professional/managerial households were more aware of these consumer-facing metrics and had resources to respond by moving to more expensive neighbhorhoods, near higher performing schools. THIS IS RATCHET EFFECT STUFF. I CUT IT BECAUSE YOU HAVEN'T INTRODUCED THE IDEA OF RATCHET EFFECT
Schools that performed well on initial league tables tended to be schools with a high share of professional/managerial families. Therefore, quantification takes a snapshot of existing social stratification and that snapshot becomes a power unto itself, amplifying the effect of previous strctural inequality on subsequent structural inequality.
```{=html}
<!-- THIS IS THE CUT TRANSITION PARAGRAPH
In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].

@@

```{=html}
<!-- SAT test-takers were categorized into one of four different market segments -- local, in-state, regional, and national -- based on SAT score-sending behavior. For a given student, each college that receives a score from the student can be defined as "local" (college located in the the same Geomarket as the student), "in-state" (same state but different Geomarket), "regional" (same region but different state), or "national" (different region). In turn, a test-taker is categorized in the "local" market segment if they submit more SAT scores to local institutions than they do to in-state, regional, or national institutions. An "in-state" student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions, etc. 
A "regional" student submits more SAT scores to regional institutions than they do to local, in-state, or national institutions. A "national" student submits more SAT scores to national institutions than they do to local, in-state, or regional institutions.
-->
```
The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family in come greater than \$35,000, percent with both parents having a BA, etc.[^5] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.[^6] Taken together, the Market Segment Profile shows a college which Geomarkets possess attractive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

@@

Quantification has reactivity, discipline, and authority effects to the extent that stakeholders care about the numbers [@RN5006]. Reactivity is the idea that salient quantitative measures cause people and organizations to change their behavior. Market research suggests that EPS was highly salient. @RN5021 reports that in 1995, 37% of 4-year publics and 49% of 4-year privates used EPS, while 41% of 4-year publics and 16% of 4-year privates used ACT's market analysis service product. <!--#  @RN5006 argue that quantification has effects to the extent that stakeholders care about the numbers.-->

Quantification disciplines actors to react in particular ways. EPS software may "discipline" [@RN5007] colleges to approach recruiting in a manner consistent with the Market Segment Model. Drawing from promotional literature [e.g., @RN5019; @RN1621], @RN4980, and background conversations with enrollment managers, the practical purpose of EPS was to inform the "travel schedule" of admissions recruiters. EPS promotional material and user guides encourages users to begin by identifying which Geomarkets they will recruit from.[^9] Second, EPS users decide which high schools they will visit within selected Geomarkets. Based on the principle of homophily, the Market Segment Model suggests that selective colleges should target Geomarkets with large numbers of affluent, college educated households, while low-income communities are left to local four-year and community colleges.

@@

The effects of quantification also lead to changes in decision-making authority, frequently reducing the discretion of local decision-makers. @RN4982 sought to develop a concrete, data-driven framework -- The Market Segment Model -- that replicates the aggregate knowledge of local admissions officers. Once this knowledge was quantified and commodified onto a CD-ROM, the local expertise of admissions officers becomes less valuable. The EPS product increases the ability of a college admissions leader -- working with College Board staff or an EM consultant -- to plan recruiting efforts centrally. In background conversations, EM professionals told us that EPS software enabled colleges to plan travel without relying on admissions officers having strong local knowledge of their territories. However, EPS software often recommended visiting the same sets of affluent, high-achieving high schools that were receiving visits from other colleges. We were told that savvier, quantitatively-adept admissions offices used EPS to visit schools that EPS recommended ignoring because there would be less competition for the good students who attended these high schools.

@@

Since @RN4982 developed The Market Segment Model, Geomarkets have been central to EPS software and play a broader role in shaping how many colleges organize recruiting territories. We therefore frame our investigation of Geomarkets as a case study of quantification to explore the extent to which this geographical building block is associated with race and class. We first address (RQ1) what is the socioeconomic and racial variation between Geomarkets and how does this variation change over time. <!--# Although we cannot gain access to EPS software, by focusing on metropolitan areas that are associated with three or more Geomarkets-->

@RN4982 viewed demand for higher education as a function of class and developed Geomarkets with an eye towards identifying geographic areas that differed from one another in terms of class composition. Given the extent of class-based residential segregation in the U.S., we therefore expect substantial socioeconomic inequality between Geomarkets in large metropolitan areas.

```{=html}
<!-- 
Additionally, @RN4982 created geomarkets in service of the Market Segment Model. The Market Segment Model argues that socioeconomic status drives whether students belong to the local, in-state, regional, or national market segment. In turn, geomarkets are described by the relative numbers of local, in-state, regional, or national students, which is substantially a function of socioeconomic status. The language of @RN4982 [pp. 11-12, quoted above] suggests that geomarket borders within large metropolitan areas may have been drawn in ways that separate affluent from less affluent communities. We present the following hypothesis:
-->
```
We also expect substantial racial inequality between Geomarkets within metropolitan areas. Interestingly, the Market Segment Model is explicitly based on socioeconomic stratification, but @RN4982 do not mention race. U.S. cities are characterized by extreme historic and contemporary residential racial segregation [@RN4779]. Unless designers intentionally consider racial segregation, selection devices that categorize people based on geographic location are likely to reproduce historical race-based inequality in opportunity [@RN4975]. Second, although geomarket borders may have been drawn along class divides, a strong correlation exists between race and wealth [@RN4884]. Third, Geomarket borders may have been drawn in a way that follows the countours of racial segregation in residential housing. Examples include: the "South and South Central Los Angeles" geomarket (CA21); the "City of Oakland" geomarket (CA07), which is surrounded by the "Alameda County excluding Oakland" geomarket (CA08); and the "Wayne County Detroit" geomarket (MI01), which is surrounded by the "Detroit's Northern Suburbs" (MI02) and "Ann Arbor" (MI03) geomarkets. <!--#  Structures built upon racialized structures are racialized structures [RN4786].-->

**Geomarket Filter in Student Search Service**. <!-- By 1984, College Board Geomarkets were incorporated as a search filter in the College Board Student Search Service \[CITE\]. --> Geomarkets also function as a student list search filter for the College Board's Student Search Service. A student list contains the contact information of prospective students who meet the search filter criteria (e.g., test score, GPA) specified by the university. Student lists are the fundamental input for undergraduate recruiting campaigns because purchased names -- alongside prospects who reach out on their own -- constitute the set of prospects who receive subsequent recruiting interventions (e.g., mail, email) designed to push them toward the application and enrollment stages of the “enrollment funnel.” @RN4895 reports that 86% of public colleges and 87% of private colleges purchase student lists.[^10] Historically, dominant list vendors are the College Board and ACT, which derived student list data from test-takers, but the test-optional movement, advances in technology, and surging private equity investments have contributed to new sources of student list data [@list_biz].

@@

Because of the extent of residential segregation in the U.S., geographic borders are a commonly studied racialized input in scholarship about algorithmic bias [@RN4775; @RN4849; @RN4887]. Student list products offer many geographic filters. Some geographic filters are based on known geographic borders (e.g., zip code, county, CBSA, state). College Board also created geographic borders that subsequently became filters in the *Student Search Service*, such as Geomarkets and Geodemographic Segment filters.[^13] In their sample, @RN5012 found that 11% of student lists purchased by public research universities filtered on Geomarkets. Drawing from @RN4786, we argue that geomarkets satisfy the two criteria of racialized inputs. First, they are ostensibly race-neutral inputs in that neither @RN4982 -- nor subsequent EPS promotional material [e.g., @RN1621] -- mention race. Second, Geomarkets may be systematically correlated with race. For example, Geomarket "CA07" is "City of Oakland" and "CA08" is "Alameda Country, excluding Oakland." A university might purchase prospect contact information by filtering on CA08 in combination with additional filters, such as year of high school graduation and SAT test score. Our study therefore extends @RN5012 by examining how the racial composition of actual student list purchases changes if universities filter on particular Geomarkets. <!--# This manuscript focuses on Geomarkets. -->

@@

In both mailings and off-campus recruiting visits, @RN4982 recommend focusing on Geomarkets with large populations of students who are deciding between "institutions like your own" (p. 44). Selective colleges primarily draw from students in the "regional" and "national" segments. @RN4982 suggest that the student list purchases of selective colleges should focus on Geomarkets with large numbers of affluent, highly educated households. By contrast, the student list purchases of local state colleges and community colleges, which rely on enrollment from students in the "local" and "in-state," should focus on nearby Geomarkets with large numbers of low-SES students. Our analyses attempt to recreate the strategy recommended by @RN4982 by showing who is included versus excluded when student list purchases filter on particular Geomarkets. Considering the correlation between race and wealth in segregated America, we suggest that a class-based list-buying strategy will increase racial stratification in college access, and vice-versa. <!--#  Furthermore, we anticipate that the students of color from affluent Geomarkets will tend to have college-educated parents and be enrolled in affluent, predominantly white schools.-->
